A template for the *arxiv* style

### 2021-11-28 main.Rmd
add in examples from vignettes

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.path = "figures/",
                      out.width = "100%", 
                      out.height = "30%")
options(prompt = 'R> ', continue = '+ ')
```

```{r echo = FALSE}
library(cubble)
library(dplyr)
library(lubridate)
```

@@

```{r cubble-diagram, echo = FALSE, fig.cap="Illustration of incoming data formats for spatio-temporal data. (1) Data comes in as a single table; (2) Separate tables for spatial and temporal variables; (3) A single table with all the parameters used to query the database and a separate table for queried data; and (4) Cubical data in array or NetCDF format.", fig.align='center', out.height = "15%"}
knitr::include_graphics(here::here("figures/input-formats/input-formats.001.png"))
```

@@

```{r def-cubble, echo = FALSE, fig.align="center", out.height="30%", fig.cap = "Illustration of nested and long cubble."}
knitr::include_graphics(here::here("figures/def-cubble/def-cubble.001.png"))
```

@@

@@

Matching  - Vig 3

@@

calc_dist <- function(df_ref, df, n = 5){
  long_ref <- df_ref$long
  lat_ref <- df_ref$lat

  df %>%
    mutate(long_ref = long_ref,
           lat_ref=  lat_ref,
           dist = rnoaa::meteo_spherical_distance(lat, long,lat_ref, long_ref)) %>%
    slice_min(dist, n = n) %>%
    mutate(city = df_ref$id) %>%
    select(-long_ref, -lat_ref)
}

(stations <- do.call("rbind", purrr::map(1:nrow(latlon_ref),
                             ~calc_dist(latlon_ref[.x, ], cand))))
```

@@

```{r eval = FALSE}
sydmel_climate <- stations %>%
  rowwise() %>%
  mutate(ts = list(meteo_pull_monitors(id,
                                       date_min = "2020-01-01",
                                       date_max = "2020-12-31",
                                       var = c("PRCP", "TMAX", "TMIN")) %>%
                     select(-id))) %>%
  as_cubble(key = id, index = date, coords = c(long, lat))
```

>>>>

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.path = "figures/",
  fig.height = 10
)
options(prompt = "R> ", continue = "+ ")
```

```{r echo = FALSE}
library(cubble)
library(forcats)
library(stringr)
library(dplyr)
library(lubridate)
library(ggplot2)
library(GGally)
library(plotly)
library(crosstalk)
library(patchwork)
```

@@

```{r cubble-diagram, echo = FALSE, out.width = "100%", out.height = "30%", fig.cap="Illustration of incoming data formats for spatio-temporal data. (1) Data comes in as a single table; (2) Separate tables for spatial and temporal variables; (3) A single table with all the parameters used to query the database and a separate table for queried data; and (4) Cubical data in array or NetCDF format.", fig.align='center', out.height = "15%"}
knitr::include_graphics(here::here("figures/input-formats/input-formats.001.png"))
```

@@

```{r def-cubble, echo = FALSE, fig.align="center", out.height="30%", out.width = "100%", fig.cap = "Illustration of nested and long cubble."}
knitr::include_graphics(here::here("figures/def-cubble/def-cubble.001.png"))
```

@@

This vignette introduces how to perform spatial and temporal manipulate in a cubble with dplyr verbs. We will illustrate with `weatherdata::historical_tmax` data, which have the historical maximum temperature recorded for Australian stations with the earliest dating back to 1859.

### Spatial manipulation

`historical_tmax` is already in a nested cubble format, which is suitable for station-wise manipulation. `rnoaa` construct the station id by prefix `ASN00` to the Bureau of Meteorology (BOM) station number. In  [BOM's numbering system](http://www.bom.gov.au/climate/cdo/about/site-num.shtml), the 2nd and 3rd digit denotes the state a station is located in and this is equivalent to the 7th to 8th digit in our string. We can mutate/ filter station in a particular state based on this information and here we add a column called `state_id` and filter out the ones in New South Wales and Victoria (ranging from 46 to 90).

```{r}
tmax_nested <- weatherdata::historical_tmax %>%
  mutate(state_id = stringr::str_sub(id, 7, 8)) %>%
  filter(between(state_id, 46, 90))
```

### Temporal manipulation

There are some operations in the time dimension we would like to make:

  - Extract observations in a particular period, say, those in 1971 to 1975 and 2016 to 2020. This can be used to compare the historical and recent climate. 
  - Summarise daily records into monthly to remove sparsity

These time dimension operations can be computed in the long form and `stretch()` converts a nested cubble to a long cubble. From a long cubble, we can write the exact  same dplyr codes to complete the two tasks:

```{r}
tmax_long <- tmax_nested %>%
  stretch() %>%
  filter(lubridate::year(date) %in% c(1971:1975, 2016:2020)) %>%
  mutate(
    month = lubridate::month(date),
    group = as.factor(ifelse(
      lubridate::year(date) > 2015,
      "2016 ~ 2020",
      "1971 ~ 1975"
    ))
  ) %>%
  group_by(month, group) %>%
  summarise(tmax = mean(tmax, na.rm = TRUE))
```

### Back to spatial

A data quality issue with the `rnoaa` data is that while it records the first and last year recorded of each series,  it doesn't report the period of missingness. For example, station `ASN00047048` starts it first record in 1957, pauses for a period from 1963 to 1990, and then resumes it recording till today. Out of the `r nrow(tmax_nested)` stations in New South Wales and Victoria, 7 stations have this issue and we would like to remove those stations from the comparison.

Again, this is a station-wise operation and to convert back from the long cubble to a nested one, use `tamp()`. Here we keep the stations with 24 observations (12 months for both periods) after the monthly aggregation.

```{r}
tmax_nested2 <- tmax_long %>%
  tamp() %>%
  filter(nrow(ts) == 24)
```

In some visualisation, we may need information from both spatial and temporal dimension. One example of this is a glyph map, where spatial variables, i.e. `longitude` and `latitude`, are used to construct the major axes and temporal variables, i.e. `month` and `tmax`, are used to construct the minor axes. This requires these variables to be in the same table, rather than in different forms. In cubble, you can append spatial variables that are invariant to the `key` withe `migrate()`

```{r}
tmax_long2 <- tmax_nested2 %>%
  stretch() %>%
  migrate(latitude, longitude)
```

### Glyph map

Now its time for the glyph map!

```{r}
gly <- glyphs(
  tmax_long2,
  x_major = "longitude",
  y_major = "latitude",
  x_minor = "month",
  y_minor = "tmax",
  height = 0.6,
  width = 1
)

nsw_vic <- ozmaps::abs_ste %>%
  filter(NAME %in% c("Victoria", "New South Wales"))

plot_map(nsw_vic, fill = "transparent") +
  geom_path(data = gly, aes(gx, gy,
    group = interaction(id, group),
    color = group
  )) +
  scale_color_brewer(palette = "Dark2") + 
  theme_bw() + 
  coord_sf(xlim = c(140, 155)) + 
  theme(legend.position = "bottom")
```

## Spatial and temporal aggregation

On top of `aus_climate` data in this package, `climate_full` from package [`weatherdata`](https://github.com/huizezhang-sherry/weatherdata) has daily climate data of `r nrow(weatherdata::climate_full)` Australia stations from 2016 to 2020. This is where these stations locate in an Australia map:

This is a lot of stations to look at for one climate variable and they can't all fit into a glyph map. What we can do is to group stations into clusters and look at the aggregated series in the glyph map. In this vignette, I will introduce how to perform spatial (and temporal) aggregation using hierarchical data structure in cubble.

First let's add a new column `ll` for calculating distance matrix in the next section and aggregate the daily precipitation into weekly measure. These steps should be familiar from the vignette *Data manipulation with cubble*.

```{r}
station_nested <- weatherdata::climate_full %>%
  mutate(ll = s2::s2_lnglat(long, lat)) %>%
  stretch() %>%
  mutate(wk = lubridate::week(date)) %>%
  group_by(wk) %>%
  summarise(prcp = sum(prcp, na.rm = TRUE)) %>%
  tamp()
```

### Hierarchical data structure

Imposing a clustering structure can be thought of as building a hierarchical structure where stations are nested within clusters. As an example to illustrate here, we use a kmean clustering algorithm based on the distance matrix and specify the number of centers to be 20. More complex algorithms can also be used for more complex problem, as long as a mapping from each station id to the cluster id can be constructed. We then join this data to our station data:

```{r}
dist_raw <- scale(s2::s2_distance_matrix(station_nested$ll, station_nested$ll))

cluster_res <- tibble(
  id = station_nested$id,
  cluster = kmeans(dist_raw, centers = 20, nstart = 50)$cluster
)

station_nested <- station_nested %>%
  left_join(cluster_res)
```

One thing we hope to do with the cluster is to find the coordinates of the centroid. These are variables variant to the station but invariant to the cluster and it would be nice to have a function that structure each cluster as a row. `switch_key()` is the function that does this: it lets you to specify a new key, say `cluster` and nests all spatial variables variant to `cluster` into a column. Temporal observations from different stations while within the same cluster are bound in the nested column `ts`.

```{r}
cluster_nested <- station_nested %>%
  switch_key(cluster)
```

This structure makes it easy to compute cluster level variable, although there are a few steps to calculate the centroid coordinates: we need to find the convex hull that wraps around the cluster, make it a polygon, find the centroid of the polygon and finally, extract the x and y coordinate of each centroid:

```{r}
cluster_nested <- cluster_nested %>%
  mutate(
    chull = list(chull(.val$long, .val$lat)),
    ll_cluster = sf::st_as_sfc(
      s2::s2_make_polygon(c(.val$long[chull]),
        c(.val$lat[chull]),
        oriented = FALSE
      )
    ),
    centroid = s2::s2_centroid(ll_cluster),
    cent_long = s2::s2_x(centroid),
    cent_lat = s2::s2_y(centroid)
  )
```

After we have got `cluster_nested`, spatial and temporal data at both levels can be easily obtained. If we use `station` and `cluster` prefix to denote the two levels and `nested` and `long` for whether the data shows the spatial or temporal dimension, the relationship among the four datasets can be illustrated in the following workflow:

Start with the original `station_nested`, `stretch()` expands the `ts` column with each station (`id`) forming a group and attach variables invariant to `id` as an attribute. `switch_key()` changes the `key` from `id` to `cluster` and nests all the spatial variables that variant to `cluster`. `stretch()` `cluster_nested` will store variables that are invariant to `cluster` as a tibble in the attribute.

Now we can obtain the aggregated series as described in the workflow diagram above and construct the glyph map with `GGally::glyphs()`:

```{r}
cluster_long <- cluster_nested %>%
  stretch(ts) %>%
  group_by(wk) %>%
  summarise(prcp = sum(prcp, na.rm = TRUE)) %>%
  migrate(cent_long, cent_lat)

gly <- GGally::glyphs(cluster_long,
  x_major = "cent_long", x_minor = "wk",
  y_major = "cent_lat", y_minor = "prcp",
  height = 2, width = 4
)

state_map <- rmapshaper::ms_simplify(ozmaps::abs_ste, keep = 2e-3)
plot_map(state_map) +
  geom_text(
    data = cluster_nested,
    aes(x = cent_long, y = cent_lat, label = cluster)
  ) +
  geom_path(
    data = gly,
    aes(x = gx, y = gy, group = gid)
  )
```

We can also look at the precipitation of each individual station within the same cluster:

```{r fig.height = 5, fig.cap = "shtishisa"}
station_long <- station_nested %>%
  stretch(ts) %>%
  migrate(cluster)
station_long %>%
  ggplot(aes(x = wk, y = prcp, group = id)) +
  geom_line(alpha = .3) +
  facet_wrap(vars(cluster), scales = "free_y", ncol = 4) +
  theme_bw()
```

Lastly, there is one series on Tasmania island standing out from others, lets look at where it is:

```{r}
# this part still needs some fixing
tas_latlong <- station_nested %>%
  filter(lat < -40) %>%
  mutate(p = max(ts$prcp)) %>%
  strip_rowwise() %>%
  filter(p == max(p))

state_map <- rmapshaper::ms_simplify(ozmaps::abs_ste, keep = 2e-3)
plot_map(state_map) +
  geom_point(data = station_nested, aes(x = long, y = lat), size = 0.5) +
  geom_sf(data = cluster_nested, aes(geometry = ll_cluster), fill = "transparent") +
  geom_point(data = tas_latlong, aes(x = long, y = lat), col = "red")
```

@@

The water level data comes from [Bureau of Meteorology](http://www.bom.gov.au/metadata/catalogue/19115/ANZCW0503900528?template=full) and has a copy in `weatherdata`. Here we extract the water course level and add a column annotate this data of type `river`. For the rainfall data, we will still use the `weatherdata::climate_full`, filtering for Victorian stations in 2020 should be pretty familiar by now. Again, we first look at where these stations are on the map first:

```{r}
river <- weatherdata::water %>%
  stretch() %>%
  select(date, Water_course_level) %>%
  tamp() %>%
  mutate(type = "river")

climate <- weatherdata::climate_full %>%
  filter(between(stringr::str_sub(id, 7, 8), 76, 90)) %>%
  stretch() %>%
  filter(lubridate::year(date) == 2020) %>%
  tamp() %>%
  mutate(type = "climate")

vic_map <- rmapshaper::ms_simplify(ozmaps::abs_ste %>% filter(NAME == "Victoria"))
plot_map(vic_map) +
  geom_point(
    data = dplyr::bind_rows(river, climate),
    aes(x = long, y = lat, color = type)
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_bw()
```

### Theory

Temporal matching checks how spatially matched pairs align temporally. We use the following chart to illustrate how the temporal matching works:

```{r echo = FALSE, fig.height=3, fig.cap = "shtishisaasdf"}
set.seed(123)
dt <- tibble(
  id = factor(c(rep("A", 31), rep("a", 31)), levels = c("A", "a")),
  date = rep(1:31, 2),
  val = c(
    c(rnorm(5), 10, rnorm(7), 5, rnorm(8), 7, rnorm(8)),
    c(
      rnorm(6, sd = 0.5), 7, rnorm(7, sd = 0.5), rnorm(6, sd = 0.5),
      4, rnorm(5, sd = 0.5), 6, rnorm(4, sd = 0.5)
    )
  )
) %>% mutate(val = ifelse(val < 0, 0, val))

circle <- tibble(
  x = c(6, 14, 23, 7, 21, 27),
  y = c(10, 5, 7, 7, 4, 6),
  id = factor(c(rep("A", 3), rep("a", 3)), levels = c("A", "a")),
  xend = x + 5
)

errorbar <- bind_rows(
  circle %>% filter(id == "A"),
  circle %>% filter(id == "A") %>% mutate(id = "a")
) %>%
  mutate(id = factor(id, c("A", "a")))

fallin <- tibble(
  id = factor(rep("a", 2), c("A", "a")),
  x = c(7, 27),
  y = -0.5,
  label = "\U2714"
)

ggplot() +
  geom_line(data = dt, aes(x = date, y = val, group = id), color = "grey60") +
  geom_vline(data = circle, aes(xintercept = x), linetype = "longdash", color = "grey60") +
  geom_errorbar(data = errorbar, aes(xmin = x, xmax = xend, y = -0.5), width = 0.5) +
  geom_point(data = circle, aes(x = x, y = y), color = "brown") +
  geom_text(data = fallin, aes(x = x, y = y, label = label), color = "brown", size = 5) +
  facet_wrap(vars(id), nrow = 2) +
  scale_y_continuous(breaks = seq(0, 10, 2)) +
  scale_x_continuous(breaks = seq(1, 31, 1)) +
  theme_bw() +
  theme(panel.grid = element_blank()) +
  labs(x = "Time", y = "Value")
```

For each spatially matched pair, say `A` and `a`, we first find the largest `n` points in each series, colored in brown points here. Here we use the largest three but you can tune this number by `temporal_n_highest`. Then we construct the interval of the largest points from one series and see how many points, from the other series, fall into the intervals. The series used to construct the interval is controlled by `temporal_independent` and the window size by `temporal_window` with a default of 5.

In this illustration, we construct the interval based on series `A` and two of the three peaks from `a` falls into this interval at Time 7 and 27.

### Rainfall translates into river level

There's another mandatory argument that hasn't been introduced above: `temporal_var_to_match`. This argument controls the variable to match and it needs to appear in both the `major` and `minor` set. In the water level matching example, we match the variable `Water_course_level` from `river` to `prcp` from `climate`, hence need to manually rename one of them to match the other, here we rename `Water_course_level` to `prcp` in `river`:

```{r}
river <- river %>%
  stretch() %>%
  rename(prcp = Water_course_level) %>%
  tamp()
```

Now we use `match_sites()` to first pair the weather stations with the river gauges spatially and then apply the temporal matching on `prcp`. We will construct the interval based on peaks in `climate` since we would expect a lag effect for precipitation to flow into the river and cause a raise in river level, hence `temporal_independent = climate`. We select the 30 highest peak from the series to construct the match by setting `temporal_n_highest = 30`. This is a tuning parameter and you can start with 10% of the points of one series (here we have daily data for a year, 10% is roughly 30 points). `temporal_min_match` filters out pairs don't have enough match and to return all the pairs, set `temporal_min_match` to `0`.

```{r}
res <- match_sites(river, climate,
  temporal_var_to_match = prcp,
  temporal_independent = climate,  
  temporal_n_highest = 30,
  temporal_min_match = 15
)

res
```

The output from temporal matching is also a cubble, with additional column `.dist` and `.group` inherent from spatial matching and `n_match` for the number of matched temporal peaks. Then you can use this output to plot the location of match or to look at the series: 
```{r out.width="100%", fig.cap="..."}
p1 <- plot_map(vic_map) +
  geom_point(
    data = res,
    aes(x = long, y = lat, color = type)
  ) +
  ggrepel::geom_label_repel(
    data = res %>% filter(type == "river"),
    aes(x = long, y = lat, label = .group)
  ) +
  scale_color_brewer(palette = "Dark2") +
  ggtitle("Victoria") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

res_long <- res %>%
  stretch(ts) %>%
  migrate(.group, type) %>%
  mutate(prcp = scale(prcp)[, 1]) %>%
  switch_key(.group) %>%
  migrate(type)

p2 <- res_long %>%
  ggplot(aes(
    x = date, y = prcp,
    group = id, color = type
  )) +
  geom_line() +
  facet_wrap(vars(.group)) +
  scale_color_brewer(palette = "Dark2", guide = "none") +
  theme_bw() +
  labs(x = "week") +
  scale_x_date(date_labels = "%b")

p1 | p2
```

## Interative graphic with cubble

```{r eval = FALSE}
set.seed(123)
tmax_data <- weatherdata::climate_full %>%
  slice_sample(n = 50) %>%
  stretch() %>%
  mutate(ym = tsibble::yearmonth(date)) %>%
  group_by(ym) %>%
  summarise(
    tmax = mean(tmax, na.rm = TRUE),
    tmin = mean(tmin, na.rm = TRUE),
    prcp = sum(prcp, na.rm = TRUE)
  ) %>%
  ungroup(ym) %>%
  mutate(
    year = as.factor(year(ym)),
    month = as.factor(month(ym)),
    prcp = sqrt(prcp + 0.001)
  ) %>%
  tamp() %>%
  mutate(median_tmax = median(ts$tmax, na.rm = TRUE)) %>%
  stretch() %>%
  migrate(median_tmax, long, lat, name) %>%
  highlight_key(~id)

p1 <- tmax_data %>%
  ggplot(aes(
    x = month, y = tmax, group = interaction(year, id),
    color = median_tmax, label = name
  )) +
  geom_line() +
  facet_wrap(vars(fct_reorder(name, -median_tmax))) +
  theme(legend.position = "none")

state_map <- rmapshaper::ms_simplify(ozmaps::abs_ste, keep = 2e-3)
p2 <- tmax_data %>%
  ggplot() +
  geom_sf(data = state_map, aes(geometry = geometry)) +
  geom_point(aes(x = long, y = lat, color = median_tmax, label = name)) +
  theme_void()

bscols(
  ggplotly(p1, width = 800, height = 800, tooltip = "label") %>%
    highlight(on = "plotly_selected", off = "plotly_deselect", color = "red"),
  ggplotly(p2, width = 800, height = 800, tooltip = "label") %>%
    highlight(on = "plotly_selected", off = "plotly_deselect", color = "red")
)
```

@@

calc_dist <- function(df_ref, df, n = 5) {
  long_ref <- df_ref$long
  lat_ref <- df_ref$lat

  df %>%
    mutate(
      long_ref = long_ref,
      lat_ref = lat_ref,
      dist = rnoaa::meteo_spherical_distance(lat, long, lat_ref, long_ref)
    ) %>%
    slice_min(dist, n = n) %>%
    mutate(city = df_ref$id) %>%
    select(-long_ref, -lat_ref)
}

(stations <- do.call("rbind", purrr::map(
  1:nrow(latlon_ref),
  ~ calc_dist(latlon_ref[.x, ], cand)
)))
```

@@

```{r eval = FALSE}
sydmel_climate <- stations %>%
  rowwise() %>%
  mutate(ts = list(meteo_pull_monitors(id,
    date_min = "2020-01-01",
    date_max = "2020-12-31",
    var = c("PRCP", "TMAX", "TMIN")
  ) %>%
    select(-id))) %>%
  as_cubble(key = id, index = date, coords = c(long, lat))
```
